Hierarchical wrinkling in a confined permeable biogel

### 2015-02-25 fig_plis/article_SI.tex
Recombine article et SI pour express submission

\usepackage{xr}
\externaldocument[S-]{supplementary_plis}

@@

Here, we introduce model permeable biogels that produce hierarchical wrinkling as a result of their interplay between their gelation dynamics and the confinement conditions. In slit geometry, we slowly acidify a dispersion of caseins, a family of milk proteins. As a result, not only does the casein form a gel as expected~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002} --this is a well-known effect used in yoghurt making-- but the gel starts to produce wrinkling patterns in cascade when casein adhesion to the slit walls is turned off, Figure~\ref{fig:dynamics} and Supplementary Videos~\ref{S-vid:transmitted} and \ref{S-vid:reconstructed}. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that upon acidification the casein network spontaneously shrinks then swells at the micron level and wrinkles at the millimetre scale. We systematically investigate such swelling-induced wrinkling of a porous gel film sandwiched between two viscous layers, the whole system being confined between two glass plates.  The main goal of the present study is to unveil the physical origin of the observed patterns and to identify the selection mechanism of the wavelength.

@@

The stability of casein dispersions strongly depends on the pH. Supplementary Video~\ref{S-vid:titration} shows the titration of a casein dispersion at 4\%w by 1 molar HCl. At pH=7, the caseins are well dispersed then they flocculate between pH=5.5 to 4 and finally at pH=3 the solution becomes clearer as caseins partially redisperse. The state diagram of caseins as a function of pH is thus related to charge stabilisation. At neutral pH caseins are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseins (pH$\approx 4.6$), casein molecules loose their negative net charge and tend to stick to each other~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002}. Beyond the isoelectric point caseins acquire a positive net charge and therefore partially re-dissolve. In the inset of Figure~\ref{fig:acidification}b, we quantify this effect and measure with UV absorption spectroscopy the amount of free caseins within the 4\%w dispersion as a function of pH, see Supplementary Methods. Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction $x_\text{free}$ is 1. Upon decreasing the pH to the isolectric point the free casein fraction drops to almost 0: almost all the casein molecules are aggregated. Further decrease of the pH of the solution leads to a partial redissolution of the casein aggregates. At pH=2 the fraction of free casein is 0.35.

@@

To gain further insight into the mechanical properties of the gel we turn to microstructural observations under confocal microscopy. As shown in Supplementary Videos~\ref{S-vid:stick63} and \ref{S-vid:nostick63} and in Supplementary Figure~\ref{S-fig:nonmonotonic} the gel consists in a space spanning fractal network with a cut-off length scale $\xi$ that corresponds to the maximum pore size~\cite{VanDijk1986}. Permeability $\alpha$ is of the order of \SI{67000}{\square\nano\metre}, see Supplementary Figure~\ref{S-fig:permeability}. The fractal model fits the structure at all times but its parameters vary non-monotonically during GDL acidification. As shown in Figure~\ref{fig:acidification}d and Supplementary Figure~\ref{S-fig:nonmonotonic}, when reaching the isolectric point $\xi$ increases up to a maximum length of \SI{7}{\micro\metre} due to the solvent expulsion out of the casein network~\cite{Lucey1998} like in viscoelastic phase separation~\cite{Tanaka2000h}. This corresponds to a tighter packing of the caseins within the network strands. Such stiffening of the strands explains the observed increase of elasticity. Beyond the isoelectric point caseins partially regain solubility, the packing within the strands loosens which results in weaker, thicker strands and smaller pores as $\xi$ decreases down to \SI{4}{\micro\metre}, hence the long-term decrease of the elastic modulus.

%XXXCe paragraphe fait-il sens ici ?XXX As seen on the confocal measurements, the gel is porous and is permeable to the solvent.  This property is determinant as it allow the gel to absorb and expel the solvent and therefore change its volume when sticky boundary are turned off. To quantify this effect, we make the casein gel inside a capillary and dip vertically the capillary in an aqueous buffer~\cite{VanDijk1986}. In the Supplementary Figure~\ref{S-fig:permeability}, we observe that the aqueous buffer goes through the gel and rises in the tube due to capillary forces. Exploitation of this experiment give the permeability of the gel: \SI{67000}{\square\nano\metre}.

@@

%The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation~\cite{Bremer1989} (Fig.~\ref{fig:acidification}a). Since caseins strongly adsorb to interfaces, the resulting gel sticks to the container walls, allowing rheological measurements (Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoelectric point makes the casein aggregates swell (Supplementary Fig.~\ref{S-fig:nonmonotonic}), thus weakening the gel~\cite{Braga2006}. In a microscopy cell of slit geometry coated so that the casein cannot adsorb on the top wall, this microscopic nonmonotonic behaviour translates into little variation of the pore size $\xi$ but to solvent expulsion out of the gel phase (syn\ae{}resis) followed by global swelling (Fig.~\ref{fig:acidification}c and Supplementary Fig.~\ref{S-fig:nonmonotonic}).

@@

Figure~\ref{fig:sideview}a shows successive side views of confocal measurements in the case where both top and bottom walls are coated, adhesion turned off, a configuration that induces wrinkling, see also Supplementary Videos~\ref{S-vid:cutDarcy} and \ref{S-vid:cutPoiseuille}. Initially the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around \SI{23}{\minute}, the gel forms and simultaneously becomes thinner and more concentrated --the fluorescence signal which is proportional to the casein concentration is brighter compared to the initial situation. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expelled on both the upper and lower sides. We describe this situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$,  Figure~\ref{fig:sideview}b. Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume, Figure~\ref{fig:sideview}c, and excess surface area of the gel film, Figure~\ref{fig:sideview}d, see Supplementary Methods. As expected from the system design, the evolution of the gel volume is non-monotonic. After reaching a minimum around 35\% of its initial value at \SI{23}{\minute}, the volume increases. Shortly after,  at \SI{36}{\minute}, the planar film destabilises and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around \SI{38}{\minute}, the amplitude of the wrinkles reaches the thickness of the slit. The size of the primary pattern is now fixed, Supplementary Figure~\ref{S-fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (\SI{44}{\minute}) and finally buckles back (\SI{53}{\minute}) to form the secondary pattern (\SI{1}{\hour}21). When the excess area saturates, the dynamics stop (\SI{3}{\hour}). The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we observe the formation of higher order patterns.

@@

Therefore we consider two limit scenarios that leads to wrinkling patterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell, i.e. $H_1\rightarrow0$, see Supplementary Video~\ref{S-vid:cutDarcy}. Creating a two dimensional blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, $(p_2-p_1)/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{p_2-p_1}{h} = \frac{dA}{dt}$. Identifying $\sigma_{\perp}=p_2-p_1$ and injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasises the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time once the gel is touching both walls, see Figure~\ref{fig:dynamics}a and Supplementary Figure~\ref{S-fig:growth}. This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate, see Supplementary Methods for the precise definition.

The opposite limiting case is when the gel film lies in the middle of the cell ($H\equiv H_1=H_2$, see Supplementary Video~\ref{S-vid:cutPoiseuille}) and impermeable ($\alpha\rightarrow 0$). Here destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}=p_2-p_1$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842} and injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

@@

Darcy and Poiseuille models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ together with estimates for all the parameters involved in either model shall allow one to discriminate between them, see Supplementary Methods about such measurements. As summarised in Supplementary Table~\ref{S-tab:data} we systematically vary (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacts the gel stiffness, permeability and $\tau$. Finally, the only parameter that we have no control upon is the initial altitude $H$ of the gel before wrinkling that varies randomly. However we are able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the wavelength associated to the secondary blister is taken as half of that of the primary pattern as shown in Supplementary Figure~\ref{S-fig:generations}.

@@

3D data are collected on a Zeiss LSM510 confocal microscope using \SI{532}{\nano\meter} laser excitation. We use either a 10x (air) lens to measure the geometry of wrinkling, see Supplementary Material, or a 63x (oil) lens to observe the microstructure of the gel. Larger scope pictures, obtained either by stitching fluorescent microscopy images (Nikon Eclipse Ti) or by transmitted/reflected light macroscope (Nikon SMZ745T/Leica DMS1000) are used to measure the wavelength of the primary pattern, Supplementary Figures~\ref{S-fig:generations} and \ref{S-fig:patterns}. The secondary pattern wavelength is taken as half of the primary wavelength, according to Supplementary Figure~\ref{S-fig:generations}.
Data shown in Figure~\ref{fig:acidification}a and c are obtained with a SevenCompact pH-meter (Mettler Toledo) and an MCR 301 rheometer (Anton Paar). For every gel composition we monitor the gelation by small amplitude oscillations (strain 0.1\%, frequency \SI{1}{\hertz}) and estimate the flexural modulus $B$ from the maximum value of the shear modulus $G^\prime$ as $B=2G^\prime/(1-\nu)$ taking $\nu=0.3$ for the Poisson ratio, a value typical of a spongy network~\cite{Greaves2011}.
Permeability is measured using a protocol found in \cite{VanDijk1986}, see Supplementary Method and Supplementary Figure~\ref{S-fig:permeability}. The amount of free casein protein present in the gel during the GDL acidification process is measured by titration and UV-spectrometry (Ocean Optics) according to the protocol found in~\cite{Roefs1986} and detailed in Supplementary Method.

\bibliographystyle{naturemag3}
\bibliography{Yaourt}

@@

>>>>

@@

Here, we introduce model permeable biogels that produce hierarchical wrinkling as a result of their interplay between their gelation dynamics and the confinement conditions. In slit geometry, we slowly acidify a dispersion of caseins, a family of milk proteins. As a result, not only does the casein form a gel as expected~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002} --this is a well-known effect used in yoghurt making-- but the gel starts to produce wrinkling patterns in cascade when casein adhesion to the slit walls is turned off, Figure~\ref{fig:dynamics} and Supplementary Videos~\ref{vid:transmitted} and \ref{vid:reconstructed}. Using a combination of titration, rheology, light microscopy and confocal microscopy, we demonstrate that upon acidification the casein network spontaneously shrinks then swells at the micron level and wrinkles at the millimetre scale. We systematically investigate such swelling-induced wrinkling of a porous gel film sandwiched between two viscous layers, the whole system being confined between two glass plates.  The main goal of the present study is to unveil the physical origin of the observed patterns and to identify the selection mechanism of the wavelength.

@@

The stability of casein dispersions strongly depends on the pH. Supplementary Video~\ref{vid:titration} shows the titration of a casein dispersion at 4\%w by 1 molar HCl. At pH=7, the caseins are well dispersed then they flocculate between pH=5.5 to 4 and finally at pH=3 the solution becomes clearer as caseins partially redisperse. The state diagram of caseins as a function of pH is thus related to charge stabilisation. At neutral pH caseins are charged negatively and repel each other. As the pH decreases to the isoelectric point of the caseins (pH$\approx 4.6$), casein molecules loose their negative net charge and tend to stick to each other~\cite{Roefs1986,Bremer1989,Lucey1998,Dickinson2002}. Beyond the isoelectric point caseins acquire a positive net charge and therefore partially re-dissolve. In the inset of Figure~\ref{fig:acidification}b, we quantify this effect and measure with UV absorption spectroscopy the amount of free caseins within the 4\%w dispersion as a function of pH, see Supplementary Methods. Starting at pH=6.8 all the casein molecules are well dispersed and the free casein fraction $x_\text{free}$ is 1. Upon decreasing the pH to the isolectric point the free casein fraction drops to almost 0: almost all the casein molecules are aggregated. Further decrease of the pH of the solution leads to a partial redissolution of the casein aggregates. At pH=2 the fraction of free casein is 0.35.

@@

To gain further insight into the mechanical properties of the gel we turn to microstructural observations under confocal microscopy. As shown in Supplementary Videos~\ref{vid:stick63} and \ref{vid:nostick63} and in Supplementary Figure~\ref{fig:nonmonotonic} the gel consists in a space spanning fractal network with a cut-off length scale $\xi$ that corresponds to the maximum pore size~\cite{VanDijk1986}. Permeability $\alpha$ is of the order of \SI{67000}{\square\nano\metre}, see Supplementary Figure~\ref{fig:permeability}. The fractal model fits the structure at all times but its parameters vary non monotonously during GDL acidification. As shown in Figure~\ref{fig:acidification}d and Supplementary Figure~\ref{fig:nonmonotonic}, when reaching the isolectric point $\xi$ increases up to a maximum length of \SI{7}{\micro\metre} due to the solvent expulsion out of the casein network~\cite{Lucey1998} like in viscoelastic phase separation~\cite{Tanaka2000h}. This corresponds to a tighter packing of the caseins within the network strands. Such stiffening of the strands explains the observed increase of elasticity. Beyond the isoelectric point caseins partially regain solubility, the packing within the strands loosens which results in weaker, thicker strands and smaller pores as $\xi$ decreases down to \SI{4}{\micro\metre}, hence the long-term decrease of the elastic modulus.

%XXXCe paragraphe fait-il sens ici ?XXX As seen on the confocal measurements, the gel is porous and is permeable to the solvent.  This property is determinant as it allow the gel to absorb and expel the solvent and therefore change its volume when sticky boundary are turned off. To quantify this effect, we make the casein gel inside a capillary and dip vertically the capillary in an aqueous buffer~\cite{VanDijk1986}. In the Supplementary Figure~\ref{fig:permeability}, we observe that the aqueous buffer goes through the gel and rises in the tube due to capillary forces. Exploitation of this experiment give the permeability of the gel: \SI{67000}{\square\nano\metre}.

@@

%The slow acidification of a protein (sodium caseinate) solution down to the isoelectric pH induces gelation~\cite{Bremer1989} (Fig.~\ref{fig:acidification}a). Since caseins strongly adsorb to interfaces, the resulting gel sticks to the container walls, allowing rheological measurements (Fig.~\ref{fig:acidification}b). Further decrease of the pH beyond the isoelectric point makes the casein aggregates swell (Supplementary Fig.~\ref{fig:nonmonotonic}), thus weakening the gel~\cite{Braga2006}. In a microscopy cell of slit geometry coated so that the casein cannot adsorb on the top wall, this microscopic nonmonotonic behaviour translates into little variation of the pore size $\xi$ but to solvent expulsion out of the gel phase (syn\ae{}resis) followed by global swelling (Fig.~\ref{fig:acidification}c and Supplementary Fig.~\ref{fig:nonmonotonic}).

@@

Figure~\ref{fig:sideview}a shows successive side views of confocal measurements in the case where both top and bottom walls are coated, adhesion turned off, a configuration that induces wrinkling, see also Supplementary Videos~\ref{vid:cutDarcy} and \ref{vid:cutPoiseuille}. Initially the dispersion is homogeneous and fills the entire slit. The black margins correspond to the glass walls. Around \SI{23}{\minute}, the gel forms and simultaneously becomes thinner and more concentrated --the fluorescence signal which is proportional to the casein concentration is brighter compared to the initial situation. This effect is the fingerprint of syn\ae{}resis: as gelation proceeds, the solvent is expelled on both the upper and lower sides. We describe this situation as a porous elastic film of thickness $h$ sandwiched between two viscous layers of respective thickness $H_1$ and $H_2$ in a cell of thickness $e=H_1+h+H_2$,  Figure~\ref{fig:sideview}b. Extracting the top and bottom surface of the gel phase allows for quantitative measurements of the relative volume, Figure~\ref{fig:sideview}c, and excess surface area of the gel film, Figure~\ref{fig:sideview}d, see Supplementary Methods. As expected from the system design, the evolution of the gel volume is non-monotonous. After reaching a minimum around 35\% of its initial value at \SI{23}{\minute}, the volume increases. Shortly after,  at \SI{36}{\minute}, the planar film destabilises and starts to wrinkle in the $z$-direction creating excess area. This confirms that the instability is due to swelling. Around \SI{38}{\minute}, the amplitude of the wrinkles reaches the thickness of the slit. The size of the primary pattern is now fixed, Supplementary Figure~\ref{fig:growth}. However swelling continues, the gel flattens on both top and bottom walls (\SI{44}{\minute}) and finally buckles back (\SI{53}{\minute}) to form the secondary pattern (\SI{1}{\hour}21). When the excess area saturates, the dynamics stop (\SI{3}{\hour}). The amount of GDL controls the ability of the gel to swell: at low GDL concentrations we only observe the primary pattern, at high GDL concentrations we observe the formation of higher order patterns.

@@

Therefore we consider two limit scenarios that leads to wrinkling patterns. In a first limit, the film of gel sits without sticking to the bottom wall of the cell, i.e. $H_1\rightarrow0$, see Supplementary Video~\ref{vid:cutDarcy}. Creating a two dimensional blister requires flowing solvent of viscosity $\eta$ through the gel of permeability $\alpha$. Darcy law~\cite{Darcy1856} relates the pressure gradient, $(p_2-p_1)/h$, to the volume flux per unit area $v = \frac{\alpha}{\eta}\frac{p_2-p_1}{h} = \frac{dA}{dt}$. Identifying $\sigma_{\perp}=p_2-p_1$ and injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Darcy mode as
\begin{equation}
\lambda_D \sim h^{1/2} \alpha^{1/4} \Upsilon^{1/4}
\end{equation}
where $\Upsilon = B\tau/\eta$ is a dimensionless viscoelastic factor quantifying the relative stiffness of the film and the solvent at a characteristic time $\tau$. This expression emphasises the kinetic nature of the wavelength selection through Darcy flow. However we note that the wavelength of a generation of the pattern does not evolve with time once the gel is touching both walls, see Figure~\ref{fig:dynamics}a and Supplementary Figure~\ref{fig:growth}. This arrest may be due to bending rigidity~\cite{LeGoff2014}. Here the relevant time scale $\tau$ corresponds to the times it takes for the wrinkle amplitude to saturate, see Supplementary Methods for the precise definition.

The opposite limiting case is when the gel film lies in the middle of the cell ($H\equiv H_1=H_2$, see Supplementary Video~\ref{vid:cutPoiseuille}) and impermeable ($\alpha\rightarrow 0$). Here destabilisation over a wavelength $\lambda$ creates a lubrication ($H\ll\lambda$) flow in the viscous layers. By symmetry the transverse load across the gel $\sigma_{\perp}=p_2-p_1$ is also the pressure difference over the wavelength. Using a Poiseuille profile for the flow, $\frac{dA}{dt} \sim \sigma_{\perp} \frac{H^3}{\eta\lambda^2}$~\cite{Poiseuille1842} and injecting this expression in equation~(\ref{eq:lstar}) yields the wavelength of the Poiseuille mode as
\begin{equation}
\lambda_P \sim (hH)^{1/2} \Upsilon^{1/6}
\end{equation}
where we recognise the scaling for the wrinkles of an elastic film on a thin ($H\ll\lambda$) elastic substrate~\cite{Cerda2003} where $\Upsilon$ would be the ratio between the flexural modulus of the film and that of the substrate. Indeed within the lubrication approximation or at low Reynolds numbers one can consider a viscous film as elastic with effective Young modulus $3\eta/\tau$~\cite{Biot1957,Boudaoud2001}.

@@

Darcy and Poiseuille models yield wavelengths that are of the same order in most of the cases under study. Thus only systematic measurements of the experimental $\lambda$ together with estimates for all the parameters involved in either model shall allow one to discriminate between them, see Supplementary Methods about such measurements. As summarised in Supplementary Table~\ref{tab:data} we systematically vary (i) the thickness of the cell, impacting $h$ and the maximal amplitude, (ii) the gel composition and thus both its stiffness and its permeability, (iii) the viscosity of the solvent by adding glycerol, which also impacts the gel stiffness, permeability and $\tau$. Finally, the only parameter that we have no control upon is the initial altitude $H$ of the gel before wrinkling that varies randomly. However we are able to obtain more data at small $H/H^*$ ratios by considering not only the primary patterns but also the secondary ones. Since the saturation of the primary pattern flattens the gel on the top and bottom walls, the secondary blisters always appear in a Darcy situation ($H\ll H^*$). Here the wavelength associated to the secondary blister is taken as half of that of the primary pattern as shown in Supplementary Figure~\ref{fig:generations}.

@@

3D data are collected on a Zeiss LSM510 confocal microscope using \SI{532}{\nano\meter} laser excitation. We use either a 10x (air) lens to measure the geometry of wrinkling, see Supplementary Material, or a 63x (oil) lens to observe the microstructure of the gel. Larger scope pictures, obtained either by stitching fluorescent microscopy images (Nikon Eclipse Ti) or by transmitted/reflected light macroscope (Nikon SMZ745T/Leica DMS1000) are used to measure the wavelength of the primary pattern, Supplementary Figures~\ref{fig:generations} and \ref{fig:patterns}. The secondary pattern wavelength is taken as half of the primary wavelength, according to Supplementary Figure~\ref{fig:generations}.
Data shown in Figure~\ref{fig:acidification}a and c are obtained with a SevenCompact pH-meter (Mettler Toledo) and an MCR 301 rheometer (Anton Paar). For every gel composition we monitor the gelation by small amplitude oscillations (strain 0.1\%, frequency \SI{1}{\hertz}) and estimate the flexural modulus $B$ from the maximum value of the shear modulus $G^\prime$ as $B=2G^\prime/(1-\nu)$ taking $\nu=0.3$ for the Poisson ratio, a value typical of a spongy network~\cite{Greaves2011}.
Permeability is measured using a protocol found in \cite{VanDijk1986}, see Supplementary Method and Supplementary Figure~\ref{fig:permeability}. The amount of free casein protein present in the gel during the GDL acidification process is measured by titration and UV-spectrometry (Ocean Optics) according to the protocol found in~\cite{Roefs1986} and detailed in Supplementary Method.

\bibliographystyle{pnas2009}
\bibliography{Yaourt}

@@

\clearpage
\newpage
\setcounter{figure}{0}

\section*{Supplementary material}

\subsection*{Fraction of free casein as a function of pH}

The fraction of free casein proteins of a \SI{100}{\milli\litre}, 4\%w sodium caseinate dispersion is obtained either by titration by 1 molar HCl, see Supplementary Video~\ref{vid:titration}, or during the action of 4\%w GDL. In both cases, \SI{0.4}{\milli\litre} of the \SI{100}{\milli\litre} dispersion is sampled regularly as a function of pH, and centrifuged (centrifuge 5424, Eppendorf) in Eppendorf tubes at 14500 rpm for \SI{3}{\minute}. The supernatant is diluted and analysed with a UV absorption spectrometer (USB4000, Ocean Optics). According to~\cite{Roefs1986}, we deduce the fraction of free casein proteins from the adsorption spectrum as:
%
\begin{align}
x_\text{free} &= \frac{A(\text{pH})}{A(\text{pH}=7)},\\
\text{with }A(\text{pH}) &= A_{280}-1.706 A_{320},
\end{align}
%
where $A_{280}$ is the absorption peak at \SI{280}{\nano\metre} due to aromatic amino acids in the casein molecules plus a turbidity background whereas the absorption peak at \SI{320}{\nano\metre} is only due to turbidity.

Above the isoelectric point, the homogeneous acidification by GDL continuously creates very small protein aggregates that are difficult to centrifuge. Such suspensions are too turbid and too rapidly evolving to reliably measure $x_\text{free}$.

\section*{Microstructure measurements}

\begin{figure}
	\includegraphics{nonmonotonic}%
	\caption{Microstructure evolution of a 4\%w sodium caseinate solution acidified with 4\%w GDL in water. (a) Details of confocal images in fully adhesive conditions (top row) and with no adhesion on top (bottom row). The scale bar is \SI{10}{\micro\metre}. Arrows indicates the pore size. (b) Corresponding Fourier spectra when no adhesion on top, same times as (a). Lines are fits to equation~(\ref{eq:fractalS}). The inset shows a schematic association between real space structures and wave numbers. (c-e) Evolution of cut-off length $\xi$, susceptibility $\chi$ and fractal dimension $d$ in full adhesive (black line) and no adhesion on top (orange dashed line) situations. Symbols indicate the times of the images in (a).}%
	\label{fig:nonmonotonic}%
\end{figure}

To follow the evolution of the microstructure of the gel, we acquire slices of $(\SI{2048}{px})^2$ with pixels of \SI{98}{\nano\metre}, Supplementary Figure~\ref{fig:nonmonotonic}a and Supplementary Videos~\ref{vid:stick63} and \ref{vid:nostick63}. After Hann windowing we perform Fourier transform and radially average the resulting spectrum to obtain $I(q)$, see Supplementary Figure~\ref{fig:nonmonotonic}b that is fitted by the following fractal form~\cite{VanDijk1986}:
\begin{equation}
I(q) = \frac{2\chi\Gamma(d)}{\left(1+\left((d+1)\xi q\right)^2\right)^{d/2}}
\label{eq:fractalS}
\end{equation}
where the susceptibility $\chi$ is related to the fluorescence contrast between the network and the solvent, the cut-off length $\xi$ corresponds to the largest pore size and $d$ is the fractal dimension. $\Gamma$ is the Gamma function. Equation~(\ref{eq:fractalS}) is a good approximation to the 2D Fourier transform of a fractal pair correlation function $g(r) \sim 1+ e^{-r/\xi} r^{d-2}$.
%En 3D on peut citer \cite{Martin1987,Chen1986}

Figure~\ref{fig:acidification}d of the main text and Supplementary Figure~\ref{fig:nonmonotonic}c-e show the evolution of the fitting parameters in two different sets of boundary conditions: (i) complete adhesion of the gel on all walls, preventing syn\ae{}resis and swelling; (ii) complete adhesion except on the top wall, allowing syn\ae{}resis and swelling. The latter situation is a proxy to the situation where we observe wrinkling, since it is necessary to keep the gel into the thin focal plane of the 63x objective lens of the confocal microscope. To do so, we coat only the microscope slide with acrylamide brush. The cover slip is cleaned but not coated, thus keeping strong adhesion on the side closer to the objective lens. Neither buckling nor wrinkling is detected in such a configuration.

In situation (i), the gel keeps its topology throughout the gelation and over-acidification. Left of Supplementary Figure~\ref{fig:nonmonotonic}a shows that the arms of the gel first shrink laterally while keeping their length, and then swell again. Syn\ae{}resis and swelling are thus local processes. By contrast in situation (ii), the gel as a whole compresses while forming. Since our focal plane is at a fixed depth, we cannot follow the topology of the same subset of the network, however the pore size is clearly smaller than in situation (ii). The isoelectric gel, at minimum gel volume, is probably isotropic or only marginally anisotropic in $z$. This explains that subsequent swelling is not occurring only along $z$ but also in plane, causing buckling.

$\xi$ is also systematically measured in the same way at the end of each wrinkling experiment and reported in Supplementary Table~\ref{tab:data}. %Final fractal dimension was always close to 2 within error bars. Comparing $\chi$ across different setting of the microscope is not relevant.

\begin{figure*}%
	\includegraphics{permeability}%
	\caption{Permeability measurements. (a) Schematic representation of the experimental set up. (b-c) Evolution of the height of the interface in tube 1 relative to the final height in tube 2. The black line is the best exponential fit $Ae^{-t/\tau}$. (b) Gel is 4\%w casein, 4\%w GDL in water, $H=\SI{2.3}{\milli\metre}$ and $\tau=\SI{57}{\minute}$. (c) Same as (b) for a 50\%w glycerol--water mixture, $H=\SI{4}{\milli\metre}$ and $\tau=\SI{100}{\hour}$.}%
	\label{fig:permeability}%
\end{figure*}
\subsection*{Permeability measurements}

To measure the permeability of our gels, we adapt a method from \cite{VanDijk1986}. First, the gel of interest is formed at the lower extremity of a thin glass tube. The tube is held vertical in a water saturated atmosphere to prevent evaporation until the pH has reached the isoelectric point ($\approx\SI{1}{\hour}$ for our most common composition). Then, we dip the tube in a bath of acetic acid/acetate buffer at pH=4. The buffer has the same glycerol content as the gel solvent, so that the viscosity is constant throughout the experiment. Alongside the first tube, see Supplementary Figure~\ref{fig:permeability} a, we dip a second identical tube with no gel in it. A webcam (Logitech Webcam Pro 9000) captures the rise of the liquid in each tube. Since the liquid height in tube 2 reaches its final value quickly, we measure directly $z_\infty-z(t)$.

Darcy law through the gel of permeability $\alpha$ and height $H$ reads 
\begin{equation}
\frac{dz}{dt} = \frac{\alpha}{\eta}\frac{\Delta P}{H}
\end{equation}
where $\Delta P=\rho g(z_\infty-z(t))$ is the hydrostatic pressure drop through the gel. We use tabulated values of water-glycerol mixtures for the density $\rho$ and viscosity $\eta$ of the buffer. We can thus write
\begin{equation}
\frac{d(z-z_\infty)}{dt} = -\frac{z-z_\infty}{\tau}\text{, with\,} \tau = \frac{\eta H}{\alpha \rho g}.
\end{equation}

By fitting $z(t)-z_\infty$ to an exponential, see Supplementary Figure~\ref{fig:permeability} b-c, we obtain $\tau$ and thus $\alpha$. Note that even with $H$ of the order of a few millimetres high glycerol contents yield $\tau\approx \SI{100}{\hour}$ due to both high viscosity and low permeability. Obtaining $z_\infty$ without tube 2 would be time prohibitive.

\subsection*{Confocal images processing}

At 10x magnification and depending on cell thickness, we acquire stacks of 20 to 50 images spaced by \SI{6.69}{\micro\metre} (Nyquist sampling). Each confocal slice is $(\SI{256}{px})^2$ with pixels of \SI{4.97}{\micro\metre}. At such low resolution details of the gel microstructure are undistinguishable as $\xi$ is smaller than the pixel size for all compositions and the gel appears as a continuous medium.

We first de-noise each plane by a short-range Gaussian blur ($\sigma=\SI{2}{px}\ll\lambda$). Then, for each $(x,y)$, we focus on the z-dependent intensity profile $I(z)$. We define the position of the top surface of the gel $z_\text{top}$ as the point where $I(z)$ first crosses half of its maximum value over the given vertical line. Since this criterion is local, we do not need to correct for background intensity variation. In this way we obtain $z_\text{top}(x,y)$ with subpixel resolution ($\pm\SI{0.1}{px}=\SI{0.67}{\micro\metre}$), and $z_\text{bottom}(x,y)$ in a symmetrical way.

From these data, we obtain for each stack the volume and thickness of the gel phase, area and peak-to-peak amplitude of the mid surface. Velocity is obtained by time differentiation. Once the instability has saturated on the top and bottom walls of the cell, peaks in the histograms of $z_\text{top}$ and $z_\text{bottom}$ yield the measure of the positions of both walls, thus of the cell thickness $e$ and retrospectively of $H_1$ and $H_2$.

\section*{Characteristic time measurements}
\begin{figure}
	\includegraphics{growth}
	\caption{Growth of the instability. (a) Evolution of the gel velocity $v$ averaged over the whole surface. Inset: Evolution of the peak-to-peak amplitude $A(t)$ in semi logarithmic scales. The black line is the best exponential fit with $\tau=\SI{90}{\second}$. (b) Evolution of the bottom profile of the gel from onset of instability (yellow) to contact to both walls (brown). Note that the longitudinal size of the blister evolves in this regime. The thick black curve is the profile at much longer times and is at lower altitude due to swelling. Note that the blister foot, highlighted by the orange vertical line, has not moved since contact.}
	\label{fig:growth}
\end{figure}

In most of our samples it is possible to measure directly the peak-to-peak amplitude $A(t)$ from our confocal measurements. However when the wavelength is too large compared to the field of view, this approach breaks down since we cannot observe both the highest and lowest points of the surface. Even when valid, such a local measurement is rather noisy and we prefer to focus on the measure of the vertical velocity $v$ of the gel that can be integrated over the whole field of view.

We note that when both measurements were available (as in Supplementary Figure~\ref{fig:growth}a), the characteristic time $\tau$ obtained from the fit of the exponential growth of $A(t)$ is about $4A_\text{max}/v_\text{max}$, where $A_\text{max} = e-h$ is the amplitude when the gel comes in contact with both walls and $v_\text{max}$ is the maximum velocity of the gel. We use this approximation of $\tau$ for all our samples. This is consistent with the fact that the blister size ($\propto\lambda$) does not evolve after contact, see Supplementary Figure~\ref{fig:growth}b, thus the kinetic wavelength is set when $A(t) = A_\text{max}$.

\subsection*{Mixed Darcy--Poiseuille model}

\subsubsection*{General framework for wrinkling}
